Autonomous Oceanographic Data Collection using Excursion Sets of Vector-valued Gaussian Random Fields

### 2020-06-22 main_new.tex
updates

  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime management. Faced with limited
  resources to understand processes in the water-column, the
  combination of statistics and autonomous robotics provides new
  opportunities for experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal ocean processes by providing informative descriptions of
  spatial characteristics of ocean phenomena. Specifically, we define
  a design
  criterion based on improved characterization of the uncertainty in
  the excursions of vector-valued Gaussian random fields, and derive
  tractable expressions for the expected Bernoulli variance reduction
  in such a framework. We demonstrate how this criterion can be used
  to prioritize sampling efforts at locations that are ambiguous,
  making exploration more effective. We use simulations to study the
  properties of methods and to compare them with state-of-the-art
  approaches, followed by results from field deployments with an
  autonomous underwater vehicle (AUV) as part of a study mapping the
  boundary of a river plume. The results demonstrate the potential of
  combining statistical methods and robotic platforms to effectively
  inform and execute data-driven environmental sampling.

@@

Motivated by the challenges related to efficient data collection strategies for the vast oceans, the combination of spatial statistics, design of experiments and marine robotics is studied in this paper. The multidisciplinary efforts enable information-driven data collection in regions of high-interest regions.

@@

Monitoring the world's oceans has gained increased importance in light of the changing climate and increasing anthropogenic impact. Central to understanding the changes taking place in the upper water-column is knowledge of the bio-geophysical interaction driven by
an agglomeration of physical forcings (e.g. wind, topography,
bathymetry, tidal influences, etc.) and incipient micro-biology driven
by planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff transported into the ocean by rivers.
These often result in a range of ecosystem-related phenomena such as
blooms and plumes, with direct and indirect effects on society. \kc{we
  should look for a citation here. Check with John Ryan}
 One of the bottlenecks in the study of such phenomena lies however in the lack of observational data with sufficient resolution. Most of this \emph{undersampling} can be attributed to the large spatio-temporal variations in which ocean processes transpire, prompting the need for effective means of data collection. 
By \emph{sampling}, we refer here primarily to the design of observational strategies in the spatial domain with the aim to pursue measurements with high scientific relevance.
Models from spatial statistics and methods for design of experiments can clearly contribute in this oceanographic sampling challenge.

@@

Data collection at sea has typically been based on static buoys, floats, or ship-based methods,
with significant logistical limitations that directly impact coverage
and sampling resolution. Modern methods using satellite remote-sensing
provide large-scale coverage but have limited resolution, are limited
to sensing the surface, and are impacted by cloud cover. Numerical
ocean models similarly find it challenging to provide detail at fine
scale \citep{Lermusiaux:2006}, and also come with computation costs that can be practically limiting. The advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling in the ocean (Fig. \ref{fig:envir1}). In particular, autonomous underwater vehicles (AUVs) have advanced the state of data collection and consequently have made robotics an integral part of ocean observation \citep{das11b,Das2015,fossuminformation,fossum18b}.

@@

Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. Missions can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations, making the data collection
\emph{adaptive}.
In doing so, the space of sampling opportunities is still limited by
the so-called waypoint graph, which forms a discretization of the search domain where the AUV can navigate, but the AUV can now modify its path at each waypoint based on in-situ measurements and calculations onboard \citep{py10,Rajan12,Rajan12b}.
Full numerical ocean models based on complex differential equations cannot be run onboard the AUV with limited computational capacity, and statistical models relying on random field assumptions are relevant as means to effectively update the onboard model from in-situ data, and to guide AUV data collection trajectories.

The work presented here is primarily inspired by a case study 
pertaining to using an AUV for spatial characterization of a frontal system generated by
a river plume. Fig. \ref{fig:nidelven} shows the survey area in
Trondheim, Norway, where cold freshwater enters from the river,
creating a strong gradient in both temperature and salinity. Because
of the local topography and the Coriolis force the cold fresh water
tends to flow to the east. Depending on the variations in river discharge, tidal
effects, coastal current and wind, this boundary often gets
distorted, and knowledge about its location is highly uncertain, making deterministic planning
challenging. Here the goal is to use AUV measurements for improved description of the interface between freshwater and oceanic water. 
It is often not possible to sample the biological variables of main interest in such AUV operations, but standard instruments provide temperature and salinity measurements which serve as proxy variables for the underlying biological oceanography phenomenon. With the help of a vector-valued random field model for temperature and salinity, one can hence aim to describe the plume to a reasonable extent.
The goal of plume characterization is in this way related to that of estimating some regions of the domain, typically excursion sets (ESs), being implicitly defined by the vector-valued random field.
In our context of environmental sampling, the joint salinity and temperature
excursions of a river plume help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}. Motivating examples for ESs of multivariate processes are also abundant in other contexts, for instance in medicine, where doctors do not rely solely on a single symptom but must see several combined effects before making a diagnosis.

The questions tackled here hence pertain to the broader area of spatial data collection and experimental designs for vector-valued random field. 
%Because of limited resources to sample the large oceanographic domains, one must plan for active learning during the operation, where a relevant criterion is used to extract the most valuable designs. 
Given the operational constraints on AUV movements and the fact that surveys rely on successive measurements along a trajectory, addressing corresponding design problems calls for particular sequential strategies. 
Our main research angle in the present work is to extend sequential design strategies from the world of spatial statistics and computer experiments to the setting of both vector-valued observational data and experimental designs for feasible robotic trajectories. We leverage and extend recent progress in expected uncertainty reduction for ESs of Gaussian random fields (GRFs) in order to address this research problem. %approach will take substantial advantage of existing work in the field of excursion set estimation mainly dedicated to computer experiments in the scalar-valued case, and where space exploration can be performed without constraints regarding the distance between successive design points. 
We briefly review recent advances in targeted sequential design of experiments based on GRFs before detailing further existing literature handling related AUV sampling problems, our contributions, and outlining the rest of the paper.

@@

 While random field modeling has been one of the main topics throughout the history of spatial statistics \citep{Krige1951a,Stein1999}, even for vector-valued random field models with associated prediction approaches such as co-Kriging \citep[See, e.g.,][]{Wackernagel2003}, there has lately been a renewed interest for random field models in the context of static or sequential experimental design, be it in the context of spatial data collection \citep{Mueller2007} or in computer experiments \citep{Santner.etal2003}. As detailed in \cite{Ginsbourger2018}, GRF models have been used in particular as a basis to sequential design of computer experiments dedicated to various goals such as global optimization and set estimation \citep{Adler.Taylor2007}. Of particular relevance to our context, \cite{Bect.etal2012} focused on strategies to reduce uncertainties on volumes of excursion exceeding a prescribed threshold, while \cite{chevalier2014fast} contributed with a computationally efficient implementation of sequential strategies for this challenge. Rather than focusing on excursion volumes, approaches were investigated in \cite{French.Sain2013,Chevalier.etal2013b,Bolin.Lindgren2015,Azzimonti.etal2016} with ambitions of estimating sets themselves. Recently, sequential designs of experiments for the conservative estimation of ESs based on GRF models were presented in \citep{Azzimonti.etal}.

Surprisingly less attention has been dedicated to sequential strategies in the case of vector-valued observations. It has been long acknowledged that co-Kriging could be efficiently used for assimilating sequentially data
\citep{Vargas-Guzman1999}, but sequential strategies for estimate features of vector-valued random fields are still in their infancy. \cite{LeGratiet.etal2015} used co-Kriging-based sequential designs to multi-fidelity computer codes and \cite{Poloczek2017} used related ideas for multi-information source optimization, but not for ES like we do here. More relevant to our setting, the PhD thesis \citep[][p.82]{stroh} mentions general possibilities of stepwise uncertainty reduction strategies for ESs in the context of designing simulations, but the outputs are mainly assumed independent.

% From earlier version; to be relocated?
%Hence statistical proxy models of the environment must be used. For our purpose, we rely on Gaussian process (GP) representations of the ocean variables of interest because they are computationally convenient and yet provide enough flexibility to realistically model the spatial variability and dependence, and the correlation between multiple processes. 
\subsection{Relation to previous AUV sampling work, contributions and outline} 
%work and proposed approach in a nutshell}
%\begin{itemize}
%    \item 
%\end{itemize}
%\textcolor{blue}{We might want to re-use some of the following (from earlier version):}

Other statistical
work in the oceanographic domain include \cite{wikle2013modern}
focusing on hierarchical statistical models; \cite{sahu2008space},
studying spatio-temporal models for sea surface temperature and
salinity data; and \cite{mellucci2018oceanic} looking at the
statistical prediction of features using an underwater glider.
In this paper the focus is not on statistical modeling, but rather on statistical properties of computations for efficient data collection. In this setting, we combine novel possibilities in marine robotics with spatial statistics and experimental design to provide useful AUV sampling designs.

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Zhang2012,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\citep{bueso1998state,zidek2019monitoring}, but variance and entropy
reduction are independent of the actual data realizations under the
assumptions of GP models, so it has limited
flexibility for active adaptation of trajectories based on what is measured. 
The use of data-driven adaptive criteria
was introduced to include more targeted sampling of regions of
scientific interest in \cite{Low2009} and \cite{fossuminformation}.

The main contributions of the current paper are
\begin{itemize}
	\item Extending expected uncertainty reduction criteria to vector-valued settings. 
	\item Closed form expressions for the expected integrated Bernoulli variance (IBV) of the excursions in GRFs.
	\item Algorithms for myopic and multiple-step ahead sequential strategies for optimizing AUV sampling with respect to the mentioned criteria.
	\item Replicable experiments on synthetic test cases with accompanying code: \url{https://github.com/CedricTravelletti/MESLAS}
	\item Results of field trials running myopic strategies onboard an AUV for the characterization of a river plume.
\end{itemize}

@@

Section \ref{sec:bg_and_notation} introduces notation and co-Kriging equations of multivariate GRFs.
Section \ref{sec:set_uq} presents uncertainty quantification (UQ) techniques on ESs of GRFs, in particular the IBV and the excursion measure variance (EMV).
Section \ref{sec:eibv} turns to the  effect of new observations on EMV and IBV, and semi-analytical expected EMV and IBV over these observations are derived. 
%The resulting formulae form the backbone of the expected sequenatil uncertainty reduction strategies presented in Section \ref{sec:heuristics}.
Section \ref{Sec:UnivarEx} illustrates the concepts on a bivariate example relevant for sampling in the temperature and salinity application.

>>>>

  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime resource management. Faced with
  limited resources in understanding processes in the water-column,
  the combination of statistics and autonomous systems provide new
  opportunities for experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal ocean processes by providing informative descriptions of
  spatial characteristics of ocean phenomena. Specifically, we define
  a design criterion based on improved characterization of the
  uncertainty in the excursions of vector-valued Gaussian random
  fields, and derive tractable expressions for the expected Bernoulli
  variance reduction in such a framework. We demonstrate how this
  criterion can be used to prioritize sampling efforts at locations
  that are ambiguous, making exploration more effective. We use
  simulations to study the properties of methods and to compare them
  with state-of-the-art approaches, followed by results from field
  deployments with an autonomous underwater vehicle (AUV) as part of a
  study mapping the boundary of a river plume. The results demonstrate
  the potential of combining statistical methods and robotic platforms
  to effectively inform and execute data-driven environmental
  sampling.

@@

Motivated by the challenges related to efficient data collection
strategies for our vast oceans, we combine spatial statistics, design
of experiments and marine robotics in this work.  The
multidisciplinary efforts enable information-driven data collection in
regions of high-interest.

@@

Monitoring the world's oceans has gained increased importance in light
of the changing climate and increasing anthropogenic impact. Central
to understanding the changes taking place in the upper water-column is
knowledge of the bio-geophysical interaction driven by an
agglomeration of physical forcings (e.g. wind, topography, bathymetry,
tidal influences, etc.) and incipient micro-biology driven by
planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff transported into the ocean by rivers.  These often
result in a range of ecosystem-related phenomena such as blooms and
plumes, with direct and indirect effects on society. \kc{we should
  look for a citation here. Check with John Ryan} One of the
bottlenecks in the study of such phenomena lies however in the lack of
observational data with sufficient resolution. Most of this
\emph{undersampling} can be attributed to the large spatio-temporal
variations in which ocean processes transpire, prompting the need for
effective means of data collection.  By \emph{sampling}, we refer here
primarily to the design of observational strategies in the spatial
domain with the aim to pursue measurements with high scientific
relevance.  Spatial statistics and methods for experimental design can
clearly contribute to this sampling challenge.

@@

Data collection at sea has typically been based on static buoys,
floats, or ship-based methods, with significant logistical limitations
that directly impact coverage and sampling resolution. Modern methods
using satellite remote-sensing provide large-scale coverage but have
limited resolution, are limited to sensing the surface, and are
impacted by cloud cover. Numerical ocean models similarly find it
challenging to provide detail at fine scale \citep{Lermusiaux:2006},
and also come with computational costs that can be limiting. The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling in the ocean (Fig. \ref{fig:envir1}). In particular,
autonomous underwater vehicles (AUVs) have advanced the state of data
collection and consequently have made robotics an integral part of
ocean observation \citep{das11b,Das2015,fossuminformation,fossum18b}.

@@

Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. Missions can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations, making the
data collection \emph{adaptive}.  In doing so, the space of sampling
opportunities is still limited by a waypoint graph, which forms a
discretization of the search domain where the AUV can navigate;
however the AUV can now modify its path at each waypoint based on
in-situ measurements and calculations onboard using onboard
deliberation \citep{py10,Rajan12,Rajan12b}.  Full numerical ocean
models based on complex differential equations cannot be run onboard
the AUV with limited computational capacity, and statistical models
relying on random field assumptions are relevant as a means to
effectively update the onboard model from in-situ data, and to guide
AUV data collection trajectories.

The work presented here is primarily inspired by a case study
pertaining to using an AUV for spatial characterization of a frontal
system generated by a river plume. Fig. \ref{fig:nidelven} shows the
survey area in Trondheim, Norway, where cold freshwater enters from
the river, creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coriolis force the
cold fresh water tends to flow east. Depending on the variations in
river discharge, tidal effects, coastal current and wind, this
boundary often gets distorted, and knowledge about its location is
highly uncertain, making deterministic planning challenging. The goal
is therefore, to use AUV measurements for improved description of the
interface between fresh and oceanic waters.  It is often not possible
to sample the biological variables of fundamental interest in such AUV
operations, but off-the-shelf instruments provide temperature and
salinity measurements which serve as proxys for the underlying
biological phenomenon. With the help of a vector-valued random field
model for temperature and salinity, one can then aim to describe the
plume.  The goal of plume characterization, in this way, relates to
that of estimating some regions of the domain, typically excursion
sets (ESs), when implicitly defined by the vector-valued random field.
In our context of environmental sampling, the joint salinity and
temperature excursions of a river plume help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}. Motivating examples for ESs of
multivariate processes are also abundant in other contexts, for
instance in medicine, where doctors do not rely solely on a single
symptom but must see several combined effects before making a
diagnosis.

The questions tackled here hence pertain to the broader area of
spatial data collection and experimental designs for vector-valued
random fields.
%Because of limited resources to sample the large oceanographic domains, one must plan for active learning during the operation, where a relevant criterion is used to extract the most valuable designs. 
Given the operational constraints on AUV movements and the fact that
surveys rely on successive measurements along a trajectory, addressing
corresponding design problems calls for sequential strategies.  Our
main research angle in the present work is to extend sequential design
strategies from the world of spatial statistics and computer
experiments to the setting of both vector-valued observational data
and experimental designs for feasible robotic trajectories. We
leverage and extend recent progress in expected uncertainty reduction
for ESs of Gaussian random fields (GRFs) in order to address this
research
problem. %approach will take substantial advantage of existing work in the field of excursion set estimation mainly dedicated to computer experiments in the scalar-valued case, and where space exploration can be performed without constraints regarding the distance between successive design points.
We briefly review recent advances in targeted sequential design of
experiments based on GRFs before detailing other literature related to
AUV sampling and our contributions prior to outlining the rest of the
paper.

@@

While random field modeling has been one of the main topics throughout
the history of spatial statistics \citep{Krige1951a,Stein1999}, even
for vector-valued random field models with associated prediction
approaches such as co-Kriging \citep[See, e.g.,][]{Wackernagel2003},
there has lately been a renewed interest for random field models in
the context of static or sequential experimental design, be it in the
context of spatial data collection \citep{Mueller2007} or in
simulation experiments \citep{Santner.etal2003}. As detailed in
\cite{Ginsbourger2018}, GRF models have been used in particular as a
basis to sequential design of simulations dedicated to various goals
such as global optimization and set estimation
\citep{Adler.Taylor2007}. Of particular relevance to our context,
\cite{Bect.etal2012} focuses on strategies to reduce uncertainty in
volumes of excursion exceeding a prescribed threshold, while
\cite{chevalier2014fast} contributes a computationally efficient
implementation of sequential strategies. Rather than focusing on
excursion volumes, approaches were investigated in
\cite{French.Sain2013,Chevalier.etal2013b,Bolin.Lindgren2015,Azzimonti.etal2016}
with ambitions of estimating sets themselves. Recently, sequential
designs of experiments for the conservative estimation of ESs based on
GRF models were presented in \citep{Azzimonti.etal}.

Surprisingly less attention has been dedicated to sequential
strategies in the case of vector-valued observations. It has been long
acknowledged that co-Kriging could be used efficiently for
assimilating sequential data \citep{Vargas-Guzman1999}, but sequential
strategies for estimating features of vector-valued random fields are
still in their infancy. \cite{LeGratiet.etal2015} used co-Kriging
based sequential designs to multi-fidelity computer codes and
\cite{Poloczek2017} used related ideas for multi-information source
optimization, but not for ES's like we do here. More relevant to our
setting, the PhD thesis \citep[][p.82]{stroh} mentions general
possibilities of stepwise uncertainty reduction strategies for ES's in
the context of designing simulations, but the outputs are mainly
assumed independent.

% From earlier version; to be relocated?
%Hence statistical proxy models of the environment must be used. For our purpose, we rely on Gaussian process (GP) representations of the ocean variables of interest because they are computationally convenient and yet provide enough flexibility to realistically model the spatial variability and dependence, and the correlation between multiple processes. 
\subsection{Previous work in AUV sampling}

Other statistical work in the oceanographic domain include
\cite{wikle2013modern} focusing on hierarchical statistical models,
\cite{sahu2008space} studying spatio-temporal models for sea surface
temperature and salinity data and \cite{mellucci2018oceanic} looking
at the statistical prediction of features using an underwater glider.
In this paper the focus is not on statistical modeling, but rather on
statistical properties of computation for efficient data
collection. In this setting, we combine novel possibilities in marine
robotics with spatial statistics and experimental design to provide
useful AUV sampling designs.

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Smith2016,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\citep{bueso1998state,zidek2019monitoring}. However, variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models, so it has limited flexibility for
active adaptation of trajectories based on what is measured.  The use
of data-driven adaptive criteria was introduced to include more
targeted sampling of regions of scientific interest in \cite{Low2009}
and \cite{fossuminformation}.

The primary contributions of this work are:

\begin{itemize}
\item Extending expected uncertainty reduction criteria to
  vector-valued settings.  
\item Closed form expressions for the expected integrated Bernoulli
  variance (IBV) of the excursions in GRFs. 
\item Algorithms for myopic and multiple-step ahead sequential
  strategies for optimizing AUV sampling with respect to the mentioned
  criteria. 
\item Replicable experiments on synthetic test cases with accompanying
  code~\footnote{\url{https://github.com/CedricTravelletti/MESLAS}}
\item Results of field trials running myopic strategies onboard an AUV
  for the characterization of a river plume. 
\end{itemize}

@@

Section \ref{sec:bg_and_notation} introduces notation and co-Kriging
equations of multivariate GRFs.  Section \ref{sec:set_uq} presents
uncertainty quantification (UQ) techniques on ESs of GRFs, in
particular the IBV and the excursion measure variance (EMV).  Section
\ref{sec:eibv} turns to the effect of new observations on EMV and IBV,
and semi-analytical expected EMV and IBV over these observations are
derived.
%The resulting formulae form the backbone of the expected sequenatil uncertainty reduction strategies presented in Section \ref{sec:heuristics}.
Section \ref{Sec:UnivarEx} illustrates the concepts on a bivariate
example relevant for sampling temperature and salinity in our domain.
